---
title: "SARS-CoV-2 platform run 4 analysis"
author: "Aaron"
date: "04/03/2020"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r imports}

# tidyverse
library(broom)
library(furrr) # <- parallel map (future_map, plan) (devtools for walk)
library(readxl) # <- read_xlsx
library(magrittr)
library(tidyverse)
library(ggbeeswarm) # <- geom_quasirandom

# workaround to enable multicore with new rstudio versions
options(future.fork.enable = TRUE)
plan(multicore)
set.seed(42)
```

```{r}
# process input
guess_max <- 100000

reads <-   read_csv('starcode.csv')
cond <-    read_csv('conditions.csv', guess_max=guess_max)
bc.map <-  read_csv('bc-map.csv', guess_max=guess_max)
```

```{r}
# ------------------------------------------------------------------------------------
# add explicit zeros to the data

explicit.zeros <- function(df, bc.map) {
  # take only bc_sets and targets from the current run
  # assumes df has been joined with condition sheet
  bc.map %>%
    filter(
      bc_set %in% unique(df$bc_set),
    ) %>%
    left_join(df, by = c('sequence', 'bc_set')) %>%
    replace_na(list(Count = 0))
}

# select the variables in the barcode map that vary
# (and any additional info you want to include)
bc.map.var <- bc.map %>%
  select('sequence', 'target', 'amplicon', 'bc_set')

# drop the centroid column as it's not needed
# coerce Count to integer to avoid weird scientic notation behavior in format_csv
df <- reads %>%
  select(-Centroid) %>%
  rename(sequence=barcode) %>% 
  inner_join(select(cond, Sample_ID, bc_set), by = 'Sample_ID') %>% 
  group_by(Sample_ID) %>%
  group_nest() %>%
  mutate(foo = future_map(data, ~explicit.zeros(.x, bc.map.var))) %>%
  select(-data) %>%
  unnest(foo) %>%
  inner_join(cond) %>%
  mutate(Count = as.integer(Count))

# write out joined data
df %>%
  write_csv('annotated_df.csv')
```

```{r}
# ===============================================================================
#                              PLOTS
# ===============================================================================

theme_pub <- function(base_size = 11, base_family = "") {
  # based on https://github.com/noamross/noamtools/blob/master/R/theme_nr.R
  # start with theme_bw and modify from there!
  theme_bw(base_size = base_size, base_family = base_family) +# %+replace%
    theme(
      # grid lines
      panel.grid.major.x = element_line(colour="#ECECEC", size=0.5, linetype=1),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_line(colour="#ECECEC", size=0.5, linetype=1),
      panel.background   = element_blank(),

      # axis options
      axis.ticks.y   = element_blank(),
      axis.title.x   = element_text(size=rel(2), vjust=0.25),
      axis.title.y   = element_text(size=rel(2), vjust=0.35),
      axis.text      = element_text(color="black", size=rel(1)),

      # legend options
      legend.title    = element_text(size=rel(1.5)),
      legend.key      = element_rect(fill="white"),
      legend.key.size = unit(1, "cm"),
      legend.text     = element_text(size=rel(1.5)),

      # facet options
      strip.text = element_text(size=rel(2)),
      strip.background = element_blank(),

      # title options
      plot.title = element_text(size=rel(2.25), vjust=0.25, hjust=0.5)
    )
}
theme_set(theme_pub())
```

Let's first just look at reads per well across our plates:

```{r, fig.width=12, fig.height=7}
df %>%
  count(Plate_ID, Sample_Well, bc_set, nCoV_amplicon, wt=Count, name='well_total') %>%
  mutate(
      Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
      Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(well_total))) +
  geom_raster() +
  facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
  scale_fill_viridis_c(option='plasma') +
  coord_equal()
```

N1 has more even coverage this time. Pretty, pretty, pretty even. S2 looks more variable, but about the same as run3. No row I dropouts this time! What the hell happened in run01?

Let's look at distribution of well totals to possibly set a cutoff:

```{r}
df %>%
  count(Plate_ID, Sample_Well, wt=Count, name='well_total') %>%
  ggplot(aes(x=well_total)) +
    geom_density() +
    scale_x_log10()
```

If we had to do a cutoff, 1e4 might be slightly better than a shot in the dark.

Regarding the N1 to S2 difference in depth, let's sum across the plates:

```{r, fig.width=4, fig.height=3}
df %>% 
  count(nCoV_amplicon, wt=Count, name='total_counts') %>%
  ggplot(aes(x=nCoV_amplicon, y=total_counts)) +
    geom_bar(stat='identity')
```

We are close to 1:1 this time, unlike run01.

Let's look at RPP30 reads between the plates/wells:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence, -target) %>% 
  spread(amplicon, Count) %>% 
  mutate(
      Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
      Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(RPP30))) +
    geom_raster() +
    facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
    scale_fill_viridis_c(option='plasma') +
    coord_equal()
```

Very low coverage on RPP30 in this run for some reason.

Let's move on to look at spike reads now that we have spikes for both samples:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence, -target) %>% 
  spread(amplicon, Count) %>% 
  mutate(
      Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
      Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(N1_spike))) +
    geom_raster() +
    facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
    scale_fill_viridis_c(option='plasma') +
    coord_equal()
```

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence, -target) %>% 
  spread(amplicon, Count) %>% 
  mutate(
      Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
      Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(S2_spike))) +
    geom_raster() +
    facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
    scale_fill_viridis_c(option='plasma') +
    coord_equal()
```

There is some amount of assay crossover, which can only be due to index switching on the sequencer. Master mixes for these two plates were prepped separately, and spike was only put into the matched plate (i.e. N1 spike on N1 plate, S2 on S2 plate). Libraries were pooled per plate, purified, and only then mixed. Could be something during bridge amplification or something optical.


Let's norm COVID to spike and plot it out across the range of RNA copies we added in:

```{r, fig.width=12, fig.height=7}
df %>%
  filter(nCoV_amplicon=='N1') %>% 
  select(-sequence, -target) %>% 
  spread(amplicon, Count) %>% 
  mutate(N1_to_N1_spike=(N1+1)/(N1_spike+1)) %>% 
  mutate(
      Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
      Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(N1_to_N1_spike))) +
    geom_raster() +
    facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
    scale_fill_viridis_c(option='plasma') +
    coord_equal()
```

Looks cool! Let's plot this across the nCoV range.

```{r, fig.width=14, fig.height=8}
neg_ctrls = df %>%
  filter(Twist_RNA_copies==0, ATCC_virus_copies==0, ATCC_RNA_copies==0) %>% 
  select(-sequence, -target) %>% 
  # select(Sample_Well, Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies, target, Count) %>% 
  spread(amplicon, Count)

df %>% 
  select(-sequence) %>% 
  anti_join(neg_ctrls) %>% 
  select(-target) %>% 
  spread(amplicon, Count) %>% 
  mutate(N1_to_N1_spike=(N1+1)/(N1_spike+1)) %>% 
  left_join(neg_ctrls) %>% 
  filter(nCoV_amplicon=='N1') %>% 
# , 
#          Twist_RNA_copies=as.character(as.integer(Twist_RNA_copies)),
#          Twist_RNA_copies=factor(Twist_RNA_copies, levels=c('0','1','3','9','27','82','247','741',
#                                                             '2220','6670','20000', '60000'))) %>% 
  gather(key = "nCoV_template", value = "copies", Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies) %>% 
  mutate(copies=copies+1e-3) %>% 
  ggplot(aes(x=copies, y=N1_to_N1_spike, group=copies)) +
    geom_boxplot(outlier.shape=NA) +
    geom_quasirandom(aes(color=lysate), alpha=0.2) +
    scale_x_log10() + scale_y_log10() +
    facet_grid(lysate~nCoV_template) + 
    theme(axis.text.x = element_text(angle = 90, vjust=0.3))
```

Looks promising, but not in NP?

```{r, fig.width=14, fig.height=10}
nCoV_amplicon_to_plot='S2'

neg_ctrls = df %>%
  filter(nCoV_amplicon==nCoV_amplicon_to_plot) %>% 
  select(-target, -sequence) %>% 
  spread(amplicon, Count) %>% 
  # mutate(copies=pmax(Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies)) %>% 
  # filter(Twist_RNA_copies==0 | ATCC_virus_copies==0 | ATCC_RNA_copies==0) %>% 
  gather(key = "nCoV_template", value = "copies", Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies) %>%
  group_by(Sample_ID) %>% 
  # select(-nCoV_template) %>% 
  mutate(is_neg_ctrl=max(copies)==0) %>% 
  filter(is_neg_ctrl) %>% 
  # select(-sequence, -target, -template) %>% 
  # select(Sample_Well, Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies, target, Count) %>% 
  mutate(S2_to_S2_spike=(S2+1)/(S2_spike+1))

# plot_df = 
df %>% 
  # mutate(copies=pmax(Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies)) %>% 
  # filter(copies>0) %>% 
  select(-sequence, -target) %>% 
  spread(amplicon, Count) %>% 
  mutate(S2_to_S2_spike=(S2+1)/(S2_spike+1)) %>% 
  filter(nCoV_amplicon==nCoV_amplicon_to_plot) %>% 
# , 
#          Twist_RNA_copies=as.character(as.integer(Twist_RNA_copies)),
#          Twist_RNA_copies=factor(Twist_RNA_copies, levels=c('0','1','3','9','27','82','247','741',
#                                                             '2220','6670','20000', '60000'))) %>% 
  gather(key = "nCoV_template", value = "copies", Twist_RNA_copies, ATCC_virus_copies, ATCC_RNA_copies) %>%
  group_by(Sample_ID) %>% 
  mutate(is_neg_ctrl=max(copies)==0) %>% 
  filter(!is_neg_ctrl, copies>0) %>% 
  rbind(neg_ctrls) %>% 
  mutate(copies=copies+1e-3) %>% filter(nCoV_template=='Twist_RNA_copies', copies==1e-3, lysate=='HEK293') %>% 
  # filter(copies==1e-3) %>% View()
  ggplot(aes(x=copies, y=S2_to_S2_spike, group=copies)) +
    geom_boxplot(outlier.shape=NA) +
    geom_quasirandom(alpha=0.4) +
    # scale_color_paletteer_c(option='plasma') +
    scale_x_log10() + scale_y_log10() +
    facet_grid(lysate~nCoV_template) + 
    theme(axis.text.x = element_text(angle = 90, vjust=0.3))
```
