---
title: "SARS-CoV-2 platform run 2 analysis"
author: "Aaron"
date: "04/02/2020"
output: 
  github_document:
    toc: true
    toc_depth: 2
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r imports}
# tidyverse
library(broom)
library(furrr) # <- parallel map (future_map, plan) (devtools for walk)
library(readxl) # <- read_xlsx
library(magrittr)
library(tidyverse)
library(ggbeeswarm) # <- geom_quasirandom

# workaround to enable multicore with new rstudio versions
options(future.fork.enable = TRUE)
plan(multicore)
```

```{r}
# process input
guess_max <- 100000


reads <-   read_csv('starcode.csv')
cond <-    read_csv('conditions.csv', guess_max=guess_max) 
bc.map <-  read_csv('bc-map.csv', guess_max=guess_max)
```

```{r}
# ------------------------------------------------------------------------------------
# add explicit zeros to the data

explicit.zeros <- function(df, bc.map) {
  # take only assays and targets from the current run
  # assumes df has been joined with condition sheet
  bc.map %>%
    filter(
      bc_set %in% unique(df$bc_set),
    ) %>%
    left_join(df, by = c('sequence', 'bc_set')) %>%
    replace_na(list(Count = 0))
}

# select the variables in the barcode map that vary
# (and any additional info you want to include)
bc.map.var <- bc.map %>%
  select('sequence', 'target', 'bc_set')

# drop the centroid column as it's not needed
# coerce Count to integer to avoid weird scientic notation behavior in format_csv
df <- reads %>%
  select(-Centroid) %>%
  rename(sequence=barcode) %>% 
  inner_join(select(cond, Sample_ID, bc_set), by = 'Sample_ID') %>% 
  group_by(Sample_ID) %>%
  group_nest() %>%
  mutate(foo = future_map(data, ~explicit.zeros(.x, bc.map.var))) %>%
  select(-data) %>%
  unnest(foo) %>%
  inner_join(cond) %>%
  mutate(Count = as.integer(Count))

# write out joined data
df %>%
  write_csv('annotated_df.csv')
```

```{r}
# ===============================================================================
#                              PLOTS
# ===============================================================================

theme_pub <- function(base_size = 11, base_family = "") {
  # based on https://github.com/noamross/noamtools/blob/master/R/theme_nr.R
  # start with theme_bw and modify from there!
  theme_bw(base_size = base_size, base_family = base_family) +# %+replace%
    theme(
      # grid lines
      panel.grid.major.x = element_line(colour="#ECECEC", size=0.5, linetype=1),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_line(colour="#ECECEC", size=0.5, linetype=1),
      panel.background   = element_blank(),
      
      # axis options
      axis.ticks.y   = element_blank(),
      axis.title.x   = element_text(size=rel(2), vjust=0.25),
      axis.title.y   = element_text(size=rel(2), vjust=0.35),
      axis.text      = element_text(color="black", size=rel(1)),
      
      # legend options
      legend.title    = element_text(size=rel(1.5)),
      legend.key      = element_rect(fill="white"),
      legend.key.size = unit(1, "cm"),
      legend.text     = element_text(size=rel(1.5)),
      
      # facet options
      strip.text = element_text(size=rel(2)),
      strip.background = element_blank(),
      
      # title options
      plot.title = element_text(size=rel(2.25), vjust=0.25, hjust=0.5)
    )
}
theme_set(theme_pub())
```

Let's first just look at reads per well across our plates:

```{r, fig.width=12, fig.height=7}
df %>%
  count(Plate_ID, Sample_Well, bc_set, nCoV_amplicon, wt=Count, name='well_total') %>%
  mutate(
    Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
    Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(well_total))) +
  geom_raster() +
  facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
  scale_fill_viridis_c(option='plasma') +
  coord_equal()
```

N1 has more even coverage this time. Pretty, pretty, pretty even. S2 looks more variable, but about the same as run3.

Regarding the N1 to S2 difference in depth, let's sum across the plates:

```{r, fig.width=4, fig.height=3}
df %>%
  count(nCoV_amplicon, wt=Count, name='total_counts') %>%
  ggplot(aes(x=nCoV_amplicon, y=total_counts)) +
  geom_bar(stat='identity')
```

We are close to 1:1 this time, unlike run01.

Let's look at RPP30 reads between the plates/wells:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(
    Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
    Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(RPP30))) +
  geom_raster() +
  facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
  scale_fill_viridis_c(option='plasma') +
  coord_equal()
```

These seem more even this time, which is nice. Perhaps it's the 40 cycles of PCR instead of 45. There is still evidence of winning i7 primers, showing up in blocks of 4 (these were added with the liquidator). The winners are different this time, though?

Let's move on to look at nCoV reads:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(
    Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
    Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(`SARS-CoV-2`))) +
  geom_raster() +
  facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
  scale_fill_viridis_c(option='plasma') +
  coord_equal()
```

We did highest conc at right of plate for this expt, so this makes sense.

Let's plot this as a proportion of total reads:

```{r, fig.width=12, fig.height=7}
df %>%
  add_count(Sample_Well, wt=Count, name='well_total') %>% 
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(SARS_to_total=(`SARS-CoV-2`)/well_total) %>% 
  mutate(
    Row = factor(str_sub(Sample_Well, 1, 1), levels = rev(LETTERS[1:16])),
    Col = str_sub(Sample_Well, 2)
  ) %>%
  ggplot(aes(x=Col, y=Row, fill=log10(SARS_to_total))) +
  geom_raster() +
  facet_wrap(~paste(Plate_ID, nCoV_amplicon, sep = ' - ')) +
  scale_fill_viridis_c(option='plasma') +
  coord_equal()
```

Let's plot this out across the range of RNA copies we added in:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(SARS_to_RPP30=`SARS-CoV-2`/RPP30, 
         Twist_RNA_copies=factor(floor(Twist_RNA_copies))) %>% 
  ggplot(aes(x=Twist_RNA_copies, y=SARS_to_RPP30, group=Twist_RNA_copies)) +
  geom_boxplot(outlier.shape=NA) +
  geom_quasirandom(aes(color=lysate), alpha=0.2) +
  scale_y_log10() +
  facet_wrap(~paste(nCoV_amplicon, sep = ' - '))
```

The data are less noisy than run01. Are we getting signal from S2 assay at 3 copies?? This is unreal.

Let's look at raw reads to see clearly where we're getting none:

```{r, fig.width=12, fig.height=7}
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(`SARS-CoV-2`=`SARS-CoV-2`+0.1,
         Twist_RNA_copies=factor(floor(Twist_RNA_copies))) %>% 
  ggplot(aes(x=Twist_RNA_copies, y=`SARS-CoV-2`, group=Twist_RNA_copies)) +
  geom_boxplot(outlier.shape=NA) +
  geom_quasirandom(alpha=0.1) +
  scale_y_log10(breaks=c(1e-1,1e0,1e1,1e2,1e3,1e4), labels=c('ND', '1e0', '1e1', '1e2', '1e3', '1e4')) +
  facet_wrap(~paste(nCoV_amplicon, sep = ' - '))
```

We get some number of reads for nCoV in every well.

We still have the technical spike oligo issue, so let's plot the spike reads across conditions to try to get a sense of where they are showing up:

```{r, fig.width=12, fig.height=7}
# plot spike reads for each condition
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(Twist_RNA_copies=factor(floor(Twist_RNA_copies))) %>% 
  mutate(spike=spike+0.1) %>% 
  rename(spike_reads=spike) %>% 
  ggplot(aes(x=Twist_RNA_copies, y=spike_reads, group=Twist_RNA_copies)) +
  geom_boxplot(outlier.shape=NA) +
  geom_quasirandom(alpha=0.1, groupOnX=TRUE) +
  scale_y_log10(breaks=c(0.1,1,10,100,1000,10000), labels=c('ND','1e0','1e1','1e2','1e3','1e4')) +
  facet_wrap(~paste(nCoV_amplicon, sep = ' - '))
```

Clearly it's not showing up in S2, which makes sense since the spike doesn't have priming sites for those primers. Let's drill down on N1 and look at proportions:

```{r, fig.width=12, fig.height=7}
# plot spike reads for each condition
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  filter(nCoV_amplicon=='N1') %>% 
  mutate(Twist_RNA_copies=factor(floor(Twist_RNA_copies))) %>% 
  mutate(spike_to_Twist=(spike+1)/`SARS-CoV-2`) %>% 
  ggplot(aes(x=Twist_RNA_copies, y=spike_to_Twist, group=Twist_RNA_copies)) +
  geom_boxplot(outlier.shape=NA) +
  geom_quasirandom(alpha=0.1) +
  scale_y_log10(breaks=c(0.1,1,10,100,1000,10000), labels=c('ND','1e0','1e1','1e2','1e3','1e4')) +
  facet_wrap(~paste(nCoV_amplicon, sep = ' - '))
```

Ratio is 1 between 2e3 and 6e3, agreeing with run01. Still our best estimate for the level of contamination.

Now, let's look at SARS reads across Twist RNA amounts, across our i5s:

```{r, fig.width=16, fig.height=7}
# plot Twist reads for each condition across i5s to look for instances of contamination
df %>%
  select(-sequence) %>% 
  spread(target, Count) %>% 
  mutate(
    Row = factor(str_sub(Sample_Well, 1, 1), levels = LETTERS[1:16]),
    Col = str_sub(Sample_Well, 2)
  ) %>%
  mutate(Twist_RNA_copies=factor(floor(Twist_RNA_copies))) %>% 
  mutate(index2=factor(index2, levels=c('CAGATCTT', 'AGTTGCGC', 'CTGTTACA', 'TCATGACA',
                                        'GAGGGCAT', 'AATCTTGC', 'CAGATTAC', 'CATCTGAG'
                                        # 'AAGATCTG', 'GCGCAACT', 'TGTAACAG', 'TGTCATGA',
                                        # 'ATGCCCTC', 'GCAAGATT', 'GTAATCTG', 'CTCAGATG',
  ))) %>% 
  mutate(`SARS-CoV-2`=`SARS-CoV-2`+0.1) %>% 
  ggplot(aes(x=Twist_RNA_copies, y=`SARS-CoV-2`, group=Twist_RNA_copies)) +
  geom_boxplot(outlier.shape=NA) +
  geom_quasirandom(alpha=0.2) +
  scale_y_log10(breaks=c(1e-1,1e0,1e1,1e2,1e3,1e4,1e5), labels=c('ND','1e0','1e1','1e2','1e3','1e4','1e5')) +
  facet_grid(nCoV_amplicon~index2) + 
  theme(axis.text.x = element_text(angle = 90, vjust=0.3))
```

Some i5s look better than others. Need to think more about this.
